Light and growth form interact to shape stomatal ratio among British angiosperms

### 2017-10-06 ms/ms.tex
changed plant habit terminology

There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes. First, I used PLANTATT data on perennation, woodiness, and height to classify species' growth form as herbaceous (annual, biennial, or perennial) or woody (shrub or tree). Following \cite{Muir_2015}, `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g. a species with both annual and perennial forms would be classified as a biennial here). Woody species are shrubs (plant height less than 4 m) or trees (plant height greater than 4 m). Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Peat_Fitter_1994b, Salguero-Gomez_etal_2016}. I retained phanerophytes, geophytes, chamaephytes, hemicryptophytes, and therophytes, but excluded data on hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. The two main differences between these growth form classifications are that 1) most shrubs and trees are lumped together as phanerophytes and 2) many geophytes and chamaephytes are lumped together with hemicryptophytes as perennials (Fig.~\ref{fig:compare-growth-forms}).

@@

Ellenberg light indicator values (\el) differed significantly among growth forms. Among Raunki\ae r life forms, therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}; ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$). Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~than shrubs and trees (Fig.~\ref{fig:gf-light}; ANOVA - $F_{4, 367}$ = 10.8, $P$ = 2.6 $\times10^{-8}$)

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified as annual, biennial, perennial, shrub, or tree, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-gf}). Raunki\ae r life form explained variation in stomatal ratio best (lower AIC; Table~\ref{table:srAIC}) , therefore we focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. % Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are 1.12, 0.38, and -0.5, respectively. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 1.2 $\times10^{-8}$ and 8.9 $\times10^{-7}$, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.24 versus 0.12). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.27-fold change in adaxial stomatal density versus a 1.13-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $2.5 $\times10^{-10}$). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ Raunki\ae r lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.2
  & 90.4
  & 0 \\
    \el~$\times$ growth form
  & 0.46
  & 0.07
  & 0.32
  & 12
  & -38.2
  & 100.4
  & 9.9 \\
    \el~+ Raunki\ae r lifeform
  & 0.46
  & 0.071
  & 0.32
  & 8
  & -40.2
  & 96.4
  & 6 \\
     \el~+ growth form
  & 0.51
  & 0.08
  & 0.31
  & 8
  & -43.4
  & 102.7
  & 12.3 \\
    Raunki\ae r lifeform
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 82 \\
    growth form
  & 0.35
  & 0.069
  & 0.13
  & 7
  & -82.5
  & 179.1
  & 88.6 \\
    \el
  & 0.64
  & 0.107
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.2 \\
    null
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.6
  & 221.1
  & 130.7 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}

>>>>

There is no universally adopted scientific classification scheme for plant growth form, therefore I statistically competed two widely used schemes based on plant habit and Raunki\ae r life form. First, I used PLANTATT data on perennation, woodiness, and height to classify species' growth form based on habit. I categorized herbaceous species as annual, biennial, or perennial and woody species as shrub or tree. Following \cite{Muir_2015}, `biennial' includes true biennials as well as species that have a mix of perennation forms (e.g. a species with both annual and perennial forms would be classified as a biennial here). Woody species are shrubs (plant height less than 4 m) or trees (plant height greater than 4 m). Next, I compared this scheme to PLANTATT data on Raunki\ae r life form \citep{Raunkiaer_1934}, which is another way to classify growth form in comparative ecology \citep[e.g.][]{Peat_Fitter_1994b, Salguero-Gomez_etal_2016}. I retained phanerophytes, geophytes, chamaephytes, hemicryptophytes, and therophytes, but excluded data on hyrdrophytes (14 species) because many of these species are hyperstomatous (Fig. \ref{fig:violin}) due to the fact that leaves may rest on the water's surface, selecting for stomata to be present on the upper surface only. The two main differences between these growth form classifications are that 1) most shrubs and trees are lumped together as phanerophytes and 2) many geophytes and chamaephytes are lumped together with hemicryptophytes as perennials (Fig.~\ref{fig:compare-growth-forms}).

@@

Ellenberg light indicator values (\el) differed significantly among growth forms. Among Raunki\ae r life forms, therophytes (annuals), hemicryptophytes (perennial herbs with buds near the soil surface), and chamaephytes (subshrubs) had greater \els~than phanerophytes (woody plants) and geophytes (perennial herbs with storage organs) (Fig.~\ref{fig:lf-light}; ANOVA - $F_{4, 367}$ = 18.3, $P$ = 1.1 $\times10^{-13}$). Likewise, herbaceous plants (annual, biennial, and perennials) had greater \els~than shrubs and trees (Fig.~\ref{fig:hf-light}; ANOVA - $F_{4, 367}$ = 10.8, $P$ = 2.6 $\times10^{-8}$)

@@

Overall, $\mathrm{SR_{even}}$ increased with \el, but there was a significant interaction ($\Delta\mathrm{AIC} > 2$, Table~\ref{table:srAIC}) between Raunki\ae r life form and \el~(Fig.~\ref{fig:SR-lf}). When classified based on plant habit, growth form interacted with \el less ($\Delta \textrm{AIC} = 2.4$; Fig.~\ref{fig:SR-hf}). Raunki\ae r life form explained variation in stomatal ratio better than habit (lower AIC; Table~\ref{table:srAIC}), therefore we focus hereafter on those analyses. Both life form and \el~significantly increased model fit, though \el~had a markedly larger effect on model AIC (Table~\ref{table:srAIC}). The significant interaction is caused by different slopes between life forms. Among life forms with the overall greatest \el~(therophytes, hemicryptophytes, and chamaephytes, see Fig.~\ref{fig:lf-light}), there was a strong positive relationship between \el~and $\mathrm{SR_{even}}$. Parametrically bootstrapped 95\% confidence intervals for the slope did not overlap zero (Fig.~\ref{fig:SR-lf}). The slope was weakly positive or not significantly different from zero in the most shade-adapted life forms (geophytes and phanerophytes), albeit the patterns were distinct in these groups. There were both hypostomatous ($\mathrm{SR_{even}} \approx 0$) and amphistomatous ($\mathrm{SR_{even}} \approx 1$) geophytes, but these were distributed across \el s. In contrast, phanerophytes were nearly always hypostomatous regardless of \el. % Allowing slopes to vary across life form signicantly increased model fit (lower AIC, Table~\ref{table:srAIC}).

@@

Adaxial (`upper') stomatal density contributed most to the evolutionary variation in stomatal ratio. The contributions of adaxial density, abaxial density, and their covariance are 1.14, 0.38, and -0.53, respectively. This implies that evolutionary variation in adaxial stomatal density is greater than that for stomatal ratio due to positive covariance between ab- and adaxial stomatal density.

Similarly, the phylogenetic SEM showed that changes in stomatal ratio associated with \el~can be attributed mostly to evolution of adaxial stomatal density (Fig.~\ref{fig:SD-light}). Both $\mathrm{sd_{ad}}$ and $\mathrm{sd_{ab}}$ increased with \el~($P =$ 6.1 $\times10^{-7}$ and 2.9 $\times10^{-5}$, respectively). However, the regression of \el~on $\mathrm{sd_{ad}}$ was 2.1$\times$ that of \el~on $\mathrm{sd_{ab}}$ (0.21 versus 0.1). Because stomatal densities were natural log-transformed, this implies an increase in \el~by one leads to a 1.23-fold change in adaxial stomatal density versus a 1.1-fold change in abaxial stomatal density. The SEM also showed a significant positive covariance between stomatal densities on each surface ($P = $1.7 $\times10^{-11}$). These results together imply that total stomatal density increases with \el, but the response is mediated mostly by increases in adaxial stomatal density.

@@

\begin{table}[ht]
 \setstretch{\stretchy}
  \topcaption{Interaction beween species' Ellenberg light indicator value (\el) and Raunki\ae r lifeform predict stomatal ratio ($\mathrm{SR_{even}}$). I compared phylogenetic linear models using the Akaike Information Criterion (AIC), where AIC = $2k - 2 \mathrm{log}(\mathcal{L})$. $k$ is the number of model parameters and $\mathcal{L}$ is the model likelihood. Given a set of candidate models, the difference in AIC between a model and the lowest AIC ($\Delta$AIC) indicates the relative fit of competing models. The correlation coefficient $r^2$ is another indicator of model fit. $\alpha$ and $\sigma^2$ are the return rate and diffusion parameters of the Ornstein-Uhlenbeck model of trait evolution. }
  \begin{center}
  \begin{tabular}{@{} l c c c c c c c @{}}
  \toprule
  Model: $\mathrm{SR_{even}} \sim$ &  $\alpha$ & $\sigma^2$ &  $r^2$ & $k$ &  log($\mathcal{L}$) &  AIC &  $\Delta$AIC \\
  \midrule  
    \el~$\times$ Raunki\ae r lifeform
  & 0.46
  & 0.068
  & 0.34
  & 12
  & -33.3
  & 90.6
  & 0 \\
    \el~$\times$ growth form
  & 0.46
  & 0.07
  & 0.32
  & 12
  & -38.2
  & 100.3
  & 9.8 \\
    \el~+ Raunki\ae r lifeform
  & 0.47
  & 0.072
  & 0.32
  & 8
  & -40.3
  & 96.5
  & 6 \\
     \el~+ growth form
  & 0.51
  & 0.08
  & 0.31
  & 8
  & -43.3
  & 102.7
  & 12.1 \\
    Raunki\ae r lifeform
  & 0.34
  & 0.067
  & 0.15
  & 7
  & -79.2
  & 172.4
  & 81.8 \\
    growth form
  & 0.35
  & 0.069
  & 0.13
  & 7
  & -82.5
  & 178.9
  & 88.4 \\
    \el
  & 0.64
  & 0.108
  & 0.26
  & 4
  & -59.3
  & 126.6
  & 36.1 \\
    null
  & 0.29
  & 0.067
  & 0
  & 3
  & -107.5
  & 221
  & 130.5 \\
	\bottomrule
	\end{tabular}
	\label{table:srAIC}
\end{center}
\end{table}
